Subset coercions in \Coq \\ \small{Draft - \today}

### 2006-09-13 papers/article/main.tex
JCF fixes and more

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{array}
\usepackage{alltt}
\usepackage{color}
\def\text{\mbox}
\pagestyle{plain}
%\usepackage{amsmath}
%\usepackage{amsthm}

@@

\newenvironment{code}{\begin{alltt}\vspace{-1.4em}}{\vspace{-1.4em}\end{alltt}}
% Caml2Tex

@@

\usepackage{framed}
%\usepackage{multicol}

@@

\abstract{We propose a new language for writing programs with dependent
  types which can be elaborated into partial \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a practical functional programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgements can be
  translated to the \CICfull (\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term, corresponding to proof obligations (or \PVS{}
  type-checking conditions). A prototype implementation of this process
  has been implemented and integrated with the \Coq{} environment.}

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent inhabitants of complex types, we immediately get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{tail} which returns
the rest element of a non-empty list. In \ML{}:
\begin{code}
  let tail = function hd :: tl -> tl | [] -> assert(false)
\end{code}

This function is partial, its domain is reduced to non-empty lists. In
\Coq, we would rather write the following:
\begin{code}
  Definition tail (l : list A) : option (list A) :=
    match l with
      hd :: tl => Some tl
    | [] => None
    end.
\end{code}

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, and also
attach a property to the result to express more precisely what \id{tail}
does:
\begin{code}
  Definition tail (l : list A) : (l <> []) -> 
  \{ l' : list A | exists a, a :: l' = l \}.
\end{code}
Now the definition's body requires some heavy plumbing of the code which is 
not affordable when programming. Moreover, the user is forced to give a
proof term justifying that \verb|l <> []| when calling the function.
We propose a solution to overcome these difficulties, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is separated into two phases. First, we have
a weak, \emph{decidable} type-checking
procedure which does not require proofs to be present in the code when
constructing objects of a subset type. In our new language \Russell, the
following is a well-typed term:
\begin{code}
  Definition tail ( l : list A | l <> [] ) : 
    \{ l' : list A | exists a, a :: l' = l \} := 
  match l with
    hd :: tl -> tl
  | [] -> []
  end
\end{code}

The specification shows that we are defining a partial function and
enforces a relation between input and output, using a dependent type, yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proofs in both branches. After type-checking,
there is an automatic elaboration into partial \Coq terms, which collects
obligations the user has to prove. In our example,
the proof assistant will ask the user to prove that:
\begin{enumerate}
\item The list \verb|tl| has the property \verb|exists a, a :: tl = hd :: tl|
  and
\item In the context where
  \id{l} is a non-empty list and $\texttt{l} = []$, the list $[]$ has the property
  \verb|exists a, a :: [] = []| (which should be obvious as the context is contradictory).
\end{enumerate}

@@

Instead of trying to find a general method for synthetizing proofs from
programs in the \CICfull, we have integrated a method which permits to
link strong specifications and purely algorithmical code. This method,
known as \ps{} in the \PVS{} system has been used with great success and
fits naturally with how we write specifications using the subset type \cite{progmartinlof}.
The main contribution of this paper is to show how the \ps{} method 
can be adapted in a proof assistant with proof terms, which formally 
justifies the extension in the first place.

The remainder of the article is organised as follows: in section \ref{sec:Russell} we present a
type system based on \CIC, which integrates subset ``subtyping'', and
prove decidability of type-checking. Then, we show how it relates to
\CIC{} by means of translations between
judgements of the two systems in section \ref{sec:Translation}. Next we present a prototype
implementation in section \ref{sec:Program}, and finally we give our
conclusions on this work and discuss future directions in section \ref{sec:Conclusion}.

@@

In \Coq, the distinction between informative and propositional parts of
a term is formalized by the \Prop/\Set{} sorts. In \Russell, we have
special support for propositions appearing in subset types. By
delimiting the use of propositions, we can separate code from proof.

Informally, this idea is already present in mathematics. When you have
an element of subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P~2)$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elts{2}{p})$ where $p$ is a proof of
$(P~2)$ (figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and their associated
properties.

@@

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \LeftLabel{\irule{Subset-$\eltpit$}}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \quad
    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \LeftLabel{\irule{Subset-$\eltpip$}}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{-2em}
  \caption{Subset type in \CIC}
  \label{fig:cci-subset} 
  \vspace{-1em}
\end{figure}

@@

    \vspace{\infvspace}
    \Conv\DP
  \end{center}
  \vspace{-2em}
  \caption{\CIC typing judgement}
  \label{fig:cic-decl-rules}
  \vspace{-1em}
\end{figure}

@@

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell new rules}
  \label{fig:typing-decl-rules}
  \vspace{-1em}
\end{figure}

%
\Russell{}'s type system is based on the Calculus of (Co-)Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but cause no problem for our main purpose which is
programming. We omit (co-)inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgement $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.

Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $`A s_1 s_2, (s_1, s_2, s_2) `: \PTSrules$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}, B$ products to benefit from
the subtyping mechanism which we will define later.

The reductions rules of this system are the standard $\beta$ and $\pi$
for projections. We denote by $\downarrow{x}$ the complete normal form
of $x$ and by $\hnf{x}$ its head normal form. We use a standard
judgemental equality $`G \type T \eqbp U : s$.

@@

\begin{figure}[h]
  \vspace{-2em}
  \subtdRules
  \vspace{-2em}
  \caption{\Russell{} conversion}
  \label{fig:conversion-decl-rules}
  \vspace{-1em}
\end{figure}

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgement. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance for
% \cite{journals/toplas/Castagna95}
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symmetry. However, it will
become important when we create coercions (see figure
\vref{fig:coerce-impl-rules-short}).

@@

\subsubsection{Properties}
We have proved the metatheory for this system in \Coq{}
(\url{http://www.lri.fr/~sozeau/research/russell/proof.en.html}).
 We have assumed that it is strongly
normalising (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$ (our proof of SR doesn't depend on SN).
Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we preferred to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
Strong normalisation could also be proved, perhaps more easily,
using the generic proof method developed in \cite{geuvers95}, by finding
a proper interpretation of \Russell terms into the Calculus of
Constructions.

So, apart from strong normalisation, we have shown all the usual structural
and metatheoretic properties of a dependent programming language, like
weakening, thinning and substitution, stability by context coercion etc\ldots
We stop here on the subject of theoretical properties of this first system, as
it gives no new insights for our purpose and we will focus more on the
algorithmic system's properties. Besides, the details of the
formalisation and the proof fall outside the scope of this paper.
% The essential properties of the $\subd$ judgement are 
% its reflexivity: if $`G \typed A : s$, then $`G \typed A \subd A$,
% symetry: if $`G \typed A, B : s$ then $`G \typed A \subd B "<=>" `G \typed
% B \subd A$ and transitivity by \irule{SubTrans}.

@@

%     \vspace{\infvspace}
%     \SubsetR\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell{} algorithmic typing, new rules}
  \label{fig:typing-algo-rules}
%  \vspace{-1em}
\end{figure*}

@@

    \vspace{\infvspace}
    \SubSubAs\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell{} algorithmic equivalence}
  \label{fig:subtyping-algo-rules}
  \vspace{-1em}
\end{figure*}

@@

\subsection{From \CIC to \lng{}}
We have presented a calculus based on \CIC with a stronger equivalence
but with restricted typing rules. We will now see how they relate
formally.
We can build a forgetful map from terms of \CIC to \Russell
(interpreting inductives as constants). It bears
relation to the $`e$ extraction function defined in
\cite{Werner:ProofIrrelevance}. Essentially, \Russell{} terms do not
contain or manipulate logical information attached to objects of subset
types, hence we must forget about it when translating.

@@

Practically, this means we can use almost all existing definitions in the
\Coq environment. This map is involutive and
the identity on \Russell terms as they do not contain the constructor
$\id{elt}$ or subset projections. Moreover, this erasure function will be
defined on all \Coq terms elaborated from \Russell.
Indeed, the second projection of a
subset element will only appear in the second component of a subset
element. Otherwise, we would have used the projection directly in the
\Russell{} term, where it is forbidden.
Indeed, in \PVS{} you can't manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgement $0 : \mysubset{x}{`N}{x \neq
  0}$, nor can you actually derive the latter, as it won't type-check.
If we allowed the second projection in \Russell, that would be doable
and we would have an inconsistency.
Hence we left only the subset type forming rule (figure
\vref{fig:typing-decl-rules}) in \Russell, while the introduction and
first projection of subset elements are internalized by the coercion judgement.

\section{From \lng{} to \CICq{}}
\label{sec:Translation}
We will now build an interpretation
$\ip{t}{`G}$ of \Russell terms $t$ in environment $`G$ into \CIC
terms. We will check at the end that it respects the $()^{\circ}$ operator
in the sense that if $t$ is well-typed in $`G$ then $\ip{t}{`G}^{\circ} = t$.

@@

\begin{center}
  \vspace{-1.7em}
  \UAX{}
  {$`G \typeq P : \Prop$}
  {$`G \typeq ?_{P} : P$}
  {}
  \DP
  \vspace{-1.7em}
\end{center}
We restrict it to objects of type $\Prop$ because we consider that the
informative part of the algorithm has been entirely given in the
original term.

We are ready to build the interpretation, which should have the following property:
\begin{proposition}[Interpretation correctness]
  If $`G \typea t : T$ then $\ipG{`G} \typeq \ip{t}{`G} : \ip{T}{`G}$.
\end{proposition}

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organised as follows: first we will
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we will define the
interpretation of terms which is mutually recursive with the precedent
algorithm. Finally we will present a proof sketch of the aforementioned
proposition.

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which build a coercion given two types (we denote it $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
improved readability. The judgement $\subimpl{`G}{c}{T}{U}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{T}{`G}$ to an object of type $\ip{U}{`G}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed \Coq{} term
(lemma \ref{lemma:coercion-correct}). 
% We
% could also directly use the interpreted types but that complicates the
% proof.

\begin{figure}[ht]
  \vspace{-1em}
  \subtisc
  \vspace{-2em}
  \caption{Coercion derivation}
  \label{fig:coerce-impl-rules-short}
  \vspace{-1em}
\end{figure}

Coercions are formalised as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(i.e., it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied.
Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and adapted for checking coherence 
of coercions, which we don't need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this wasn't adapted to our setting, so we invented this
new presentation.

@@

Next, we have the rules for products and sums, which compose
coercions. In the \irule{SubProd} rule, we first coerce the argument to
an object of type $\ip{T}{`G}$. Then we apply it to the coerced
function, resulting in an object $\ctxdot~(c_1[x])$ of type
$\ip{V}{`G, x : T}[c_1[x]/x]$. We'll see later that this type is equivalent to
$\ip{V}{`G, x : U}$, so we can apply the second coercion and get an
object of type $\ip{W}{`G}$. There is a little twist when coercing the
second component of a dependent sum. Indeed, we must simultaneously
substitute the first projection of the coerced sum in the second
coercion which was typed in environment $`G, x : T$ and also instantiate
the coercion with the second component. Clearly, the $\ctxdot$ of the
former shouldn't be instantiated with the later, so we must first
instantiate $c_2$ and then substitute $\pi_1~\ctxdot$ for $x$ 
($x$ cannot appear in $\pi_2~t$ where $t : \Sigma x : T.V$).

@@

\begin{figure}[ht]
  \vspace{-2em}
  \interparr
  \vspace{-2em}
  \caption{Interpretation of terms}
  \label{fig:interp}
  \vspace{-1em}
\end{figure}

>>>>

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{array}
\usepackage{alltt}
\usepackage{color}
\def\text{\mbox}
%\pagestyle{plain}
%\usepackage{amsmath}
%\usepackage{amsthm}

@@

\newenvironment{code}{\begin{alltt}}{\end{alltt}}
% Caml2Tex

@@

%\usepackage{framed}
%\usepackage{multicol}

@@

\begin{abstract}
  We propose a new language for writing programs with dependent
  types which can be elaborated into partial \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a practical functional programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgements can be
  translated to the \CICfull~(\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term, corresponding to proof obligations (or \PVS{}
  type-checking conditions). A prototype implementation of this process
  is integrated with the \Coq{} environment.
\end{abstract}

\section{Introduction}
There are many means to program in the \Coq{} environment \cite{CoqArt}. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent inhabitants of complex types, we immediately get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{tl} which returns
the tail of a non-empty list. In \ML{}:
\begin{code}
  let tl = function hd :: tl -> tl | [] -> assert(false)
\end{code}

This function is partial, its domain is reduced to non-empty lists. In
\Coq, we would rather write the following:
\begin{code}
  Definition tl (l : list A) : option (list A) :=
    match l with
      hd :: tl => Some tl
    | [] => None
    end.
\end{code}

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, and also
attach a property to the result to express more precisely what \id{tl}
does:
\begin{code}
  Definition tl (l : list A) : (l <> []) -> 
  \{ l' : list A | exists a, a :: l' = l \}.
\end{code}
Now the definition's body requires some heavy plumbing of the code which is 
not affordable when programming. Moreover, the user is forced to give a
proof term justifying that \verb|l <> []| when calling the function.
We propose a solution to overcome these difficulties, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is separated into two phases. First, we have
a weak, \emph{decidable} type-checking
procedure which does not require proofs to be present in the code when
constructing objects of a subset type. In our new language \Russell, the
following is a well-typed term:
\begin{code}
  Program Definition tl ( l : list A | l <> [] ) : 
    \{ l' : list A | exists a, a :: l' = l \} := 
  match l with
    hd :: tl -> tl
  | [] -> []
  end
\end{code}

The specification shows that we are defining a partial function and
enforces a relation between input and output, using a dependent type, yet the
code remains as simple as the \ML{} definition. This is only possible
because we do not require the user to write proofs in the code. After type-checking,
there is an automatic elaboration into partial \Coq terms, which collects
obligations the user has to prove. In our example,
the proof assistant will ask the user to prove that:
\begin{enumerate}
\item The list \verb|tl| has the property \verb|exists a, a :: tl = hd :: tl|,
  and
\item In the context where
  \id{l} is a non-empty list and $\texttt{l} = []$, the list $[]$ has the property
  \verb|exists a, a :: [] = []| (which should be obvious as the context is contradictory).
\end{enumerate}

@@

Instead of trying to find a general method for synthetizing proofs from
programs in the \CICfull, we have integrated a method which permits to
link strong specifications and purely algorithmical code. This method,
known as \ps{} in the \PVS{} system, has been used with great success and
fits naturally with how we write specifications using the subset type \cite{progmartinlof}.
The main contribution of this paper is to show how the \ps{} method 
can be adapted in a proof assistant with proof terms, which formally 
justifies the extension in the first place.

The remaining of the article is organised as follows: in section \ref{sec:Russell} we present a
type system based on \CIC, which integrates subset ``subtyping'', and
prove decidability of type-checking. Then, we show how it relates to
\CIC{} by means of translations between
judgements of the two systems in section \ref{sec:Translation}. Next we present a prototype
implementation in section \ref{sec:Program}, and finally we give our
conclusions on this work and discuss future directions in section \ref{sec:Conclusion}.

@@

In \Coq, the distinction between informative and propositional parts of
a term is formalised by the \Prop/\Set{} sorts. In \Russell, we have
special support for propositions appearing in subset types. By
delimiting the use of propositions, we can separate code from proof.

Informally, this idea is already present in mathematics. When you have
an element of subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P~2)$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elt{\nat}{P}{2}{p})$ where $p$ is a proof of
$(P~2)$ (Figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and their associated
properties.

@@

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \LeftLabel{\irule{Subset-$\eltpit$}}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \quad
    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \LeftLabel{\irule{Subset-$\eltpip$}}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{-1em}
  \caption{Subset type in \CIC}
  \label{fig:cci-subset} 
  \vspace{-1em}
\end{figure}

@@

    \vspace{\infvspace}
    \Conv\DP
  \end{center}
  \vspace{-1em}
  \caption{\CIC typing judgement}
  \label{fig:cic-decl-rules}
  \vspace{-1em}
\end{figure}

@@

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{-1em}
  \caption{\Russell new rules}
  \label{fig:typing-decl-rules}
  \vspace{-1em}
\end{figure}

%
\Russell{}'s type system is based on the Calculus of Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but causes no problem for our main purpose which is
programming. We omit inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgement $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.

Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $`A s_1 s_2, (s_1, s_2, s_2) `: \PTSrules$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}, B$ to benefit from
the subtyping mechanism which we will define later.

The reduction rules of this system are the standard $\beta$ and $\pi$
for projections. We denote by $\downarrow{x}$ the complete normal form
of $x$ and by $\hnf{x}$ its head normal form. We use a standard
judgemental equality $`G \type T \eqbp U : s$.

@@

\begin{figure}[h]
  \vspace{-1em}
  \subtdRules
  \vspace{-1em}
  \caption{\Russell{} conversion}
  \label{fig:conversion-decl-rules}
  \vspace{-1em}
\end{figure}

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgement. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgement builds an \emph{equivalence} and has
proper modularity. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance for
% \cite{journals/toplas/Castagna95}
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symmetry. However, it will
become important when we create coercions (see figure
\vref{fig:coerce-impl-rules-short}).

@@

\subsubsection{Properties}
We have proved the metatheory for this system in \Coq{}
(\url{http://www.lri.fr/~sozeau/research/russell/proof.en.html}).
 We have assumed that it is strongly
normalising (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$ (our proof of SR does not depend on SN).
Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we preferred to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
Strong normalisation could also be proved, perhaps more easily,
using the generic proof method developed by Geuver \cite{geuvers95}, by finding
a proper interpretation of \Russell terms into the Calculus of
Constructions.

So, apart from strong normalisation, we have shown all the usual structural
and metatheoretic properties of a dependent programming language, like
weakening, thinning and substitution, stability by context coercion, etc.
We stop here on the subject of theoretical properties of this first system, as
it gives no new insights for our purpose and we will focus more on the
algorithmic system's properties. Besides, the details of the
formalisation and the proof fall outside the scope of this paper.
% The essential properties of the $\subd$ judgement are 
% its reflexivity: if $`G \typed A : s$, then $`G \typed A \subd A$,
% symetry: if $`G \typed A, B : s$ then $`G \typed A \subd B "<=>" `G \typed
% B \subd A$ and transitivity by \irule{SubTrans}.

@@

%     \vspace{\infvspace}
%     \SubsetR\DP
  \end{center}
  \vspace{-1em}
  \caption{\Russell{} algorithmic typing, new rules}
  \label{fig:typing-algo-rules}
%  \vspace{-1em}
\end{figure*}

@@

    \vspace{\infvspace}
    \SubSubAs\DP
  \end{center}
  \vspace{-1em}
  \caption{\Russell{} algorithmic equivalence}
  \label{fig:subtyping-algo-rules}
  \vspace{-1em}
\end{figure*}

@@

\subsection{From \CIC to \lng{}}
We have presented a calculus based on \CIC with a stronger equivalence
but with restricted typing rules. We will now see how they relate
formally.
We can build a forgetful map from terms of \CIC to \Russell
(interpreting inductives as constants). It bears
relation to the $`e$ extraction function defined by Werner in
\cite{Werner:ProofIrrelevance}. Essentially, \Russell{} terms do not
contain or manipulate logical information attached to objects of subset
types, hence we must forget about it when translating.

@@

Practically, this means we can use almost all existing definitions in the
\Coq environment. This map is involutive and
the identity on \Russell terms as they do not contain the constructor
$\id{elt}$ or subset projections. Moreover, this erasure function will be
defined on all \Coq terms elaborated from \Russell.
Indeed, the second projection of a
subset element will only appear in the second component of a subset
element. Otherwise, we would have used the projection directly in the
\Russell{} term, where it is forbidden.
Indeed, in \PVS{} you cannot manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgement $0 : \mysubset{x}{`N}{x \neq
  0}$, nor can you actually derive the latter, as it will not type-check.
If we allowed the second projection in \Russell, that would be doable
and we would have an inconsistency.
Hence we left only the subset type forming rule (figure
\vref{fig:typing-decl-rules}) in \Russell, while the introduction and
first projection of subset elements are internalized by the coercion judgement.

\section{From \lng{} to \CICq{}}
\label{sec:Translation}
We now build an interpretation
$\ip{t}{`G}$ of \Russell terms $t$ in environment $`G$ into \CIC
terms. We will check at the end that it respects the $()^{\circ}$ operator
in the sense that if $t$ is well-typed in $`G$ then $\ip{t}{`G}^{\circ} = t$.

@@

\begin{center}
  \vspace{-1em}
  \UAX{}
  {$`G \typeq P : \Prop$}
  {$`G \typeq ?_{P} : P$}
  {}
  \DP
  \vspace{-1em}
\end{center}
We restrict it to objects of type $\Prop$ because we consider that the
informative part of the algorithm has been entirely given in the
original term.

We are ready to build the interpretation, which should have the following property:
\begin{proposition}[Interpretation correctness]
  \label{interp-correct}
  If $`G \typea t : T$ then $\ipG{`G} \typeq \ip{t}{`G} : \ip{T}{`G}$.
\end{proposition}

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organised as follows: first we 
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we define the
interpretation of terms which is mutually recursive with the previous
algorithm. Finally we present a proof sketch of the aforementioned
proposition.

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which builds a coercion given two types (we denote it $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
better readability. The judgement $\subimpl{`G}{c}{T}{U}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{T}{`G}$ to an object of type $\ip{U}{`G}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed \Coq{} term
(lemma \ref{lemma:coercion-correct}). 
% We
% could also directly use the interpreted types but that complicates the
% proof.

\begin{figure}[ht]
  \vspace{-1em}
  \subtisc
  \vspace{-1em}
  \caption{Coercion derivation}
  \label{fig:coerce-impl-rules-short}
  \vspace{-1em}
\end{figure}

Coercions are formalised as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(i.e., it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied.
Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and adapted for checking coherence 
of coercions, which we do not need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this was not adapted to our setting, so we invented this
new presentation.

@@

Next, we have the rules for products and sums, which compose
coercions. In the \irule{SubProd} rule, we first coerce the argument to
an object of type $\ip{T}{`G}$. Then we apply it to the coerced
function, resulting in an object $\ctxdot~(c_1[x])$ of type
$\ip{V}{`G, x : T}[c_1[x]/x]$. We will see later that this type is equivalent to
$\ip{V}{`G, x : U}$, so we can apply the second coercion and get an
object of type $\ip{W}{`G}$. There is a little twist when coercing the
second component of a dependent sum. Indeed, we must simultaneously
substitute the first projection of the coerced sum in the second
coercion which was typed in environment $`G, x : T$ and also instantiate
the coercion with the second component. Clearly, the $\ctxdot$ of the
former shouldn't be instantiated with the later, so we must first
instantiate $c_2$ and then substitute $\pi_1~\ctxdot$ for $x$ 
($x$ cannot appear in $\pi_2~t$ where $t : \Sigma x : T.V$).

@@

\begin{figure}[ht]
  \vspace{-1em}
  \interparr
  \vspace{-1em}
  \caption{Interpretation of terms}
  \label{fig:interp}
  \vspace{-1em}
\end{figure}
